Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-30 paper/paper.tex
reviews on 3oth

%% We use the  result of the analysis to annotate  each potential garbage
%% collection  point  in   the  program  with  a   set  of  deterministic
%% finite-state automata (DFA) describing the liveness at that point.
To show the effectiveness of our method, we
implemented a copying collector that uses the results of the liveness
analysis to preserve live 
objects, both evaluated (i.e., in WHNF) and closures.  Our experiments
confirm  that  for  programs  running with  a  liveness-based  garbage
collector,   there  is   a   significant  decrease   in  peak   memory
requirements.   In  addition,   a  sizable  reduction  in   the  number  of
collections  ensures that  in spite  of using  a more  complex garbage
collector,  the execution  time  of  a program  remains comparable  with
the same program running with a reachabilty-based collector.
\warning{On our benchmark programs, we obtain a decrease of up to 75\%
  in the number of collections, and up to 50\% decrease in peak memory
  requirements.}
\end{abstract}

@@

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The  analysis  has  two parts.   We  first  calculate  a
context-sensitive  summary   of  each   function  as  a   {\em  demand
  transformer} that transforms a {\em  symbolic} demand on its body to
demands  on its  arguments.   This  summary is  used  to step  through
function  calls  during analysis.   The  concrete demand  on  a
function body is obtained through a conservative approximation similar
to  0-CFA~\cite{Shivers:1988} that  combines  the demands  on all  the
calls to the function.  The result of the analysis is an annotation of
certain program points with  deterministic finite-state automata (DFA)
capturing the liveness of variables at these points.  Depending on the
where  garbage collection  has been  triggered, the
collector
consults a  set of automata  to restrict reachability  during marking.
This results in an increase  in the garbage reclaimed and consequently
in fewer garbage collections.

While this idea has been shown  to be effective for a first-order {\em
  eager}  language~\cite{asati14lgc}, a  straightforward extension  of
the technique is not possible for lazy languages, where heap-allocated
objects may  include closures (runtime representations  of unevaluated
expressions).   Firstly, since  data  is made  live  by evaluation  of
closures, and  in lazy languages the  place in the program  where this
evaluation  takes  place  cannot be  statically  determined,  laziness
complicates  liveness analysis  itself.  Moreover,  for liveness-based
garbage collection to  be effective, we need to extend  it to closures
apart from  evaluated data.  Since a  closure can escape the  scope in
which it was created, it has  to carry the liveness information of its
free variables. Moreover, as  execution progresses and possible future
uses are eliminated,  we update the liveness information  in a closure
with a  more precise version.  Because of these reasons,   the garbage
collector  also   becomes  significantly   more  complicated   than  a
liveness-based collector for a eager language.

\warning{Experiments  with  a   single  generation  copying  collector
  (Section~\ref{sec:experiments})  confirm  the  expected  performance
  benefits.   Liveness-based  collection  results in  an  increase  in
  garbage reclaimed.   As a  consequence, there is  a reduction  in the
  number of collections up to ???\% and a decrease in the
  minimum memory requirement up to ???\%. 
  As an added benefit, there is also a reduction in the overall
  execution time for some programs.}

@@

$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, \heap,\, (+~x~y)
\rightsquigarrow
\rho',\, (\rho, \rho(x), (+~x~y))\!:\!S,\, \heap,\, s$      &
{\sc prim-1-clo} \\
\hline
$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho,\, S,\, \heap,\, (+~x~y)
\rightsquigarrow
\rho', (\rho,\, \rho(y),\, (+~x~y))\!:\!S,\, \heap,\, s$      &
{\sc prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho,\, S,\, \heap,\,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})],\, S,\, \heap,\, e_{\mathit{f}}$      &
{\sc funcall} \\
\hline
$\ell$ is a new location& {$\rho,\, S,\, \heap,\, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell],\, S,\, \heap[\ell \mapsto \langle s,
    \lfloor\rho\rfloor_{FV(s)} %% \oplus [x \mapsto  \ell] -- NO LAZY LETS
    \rangle],\, e$} &
{\sc let} \\
\hline
$\heap(\rho(x)) \ne 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho,\, S,\, \heap,\,  e_1$ & {\sc if-true} \\
\hline
$\heap(\rho(x)) = 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho,\, S, \,\heap, \, e_2$ & {\sc if-false} \\
\hline
$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho',\, (\rho, \rho(x), (\SIF~x~e_1~e_2))\!:\!S,\, \heap, \, s$}
&
{\sc if-clo} \\
\hline
{$\heap(\rho(x))~\mbox{is in WHNF with value}~v$}& $\rho,\, (\rho', \ell,
e)\!:\!S,\, \heap,\,
(\SRETURN~x)  \rightsquigarrow \rho',\, S,\, \heap[\ell := v],\, e$ &
{\sc return-whnf}\\
\hline
$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\, (\SRETURN~x)
  \rightsquigarrow$
$\rho',\, (\rho, \rho(x), (\SRETURN~x))\!:\!S,\, \heap,\,  s$} &
{\sc return-clo} \\
\hline
\end{tabular}
\caption{The small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
%}\comment
\vspace*{-6pt}
\end{figure*}

\subsection{Semantics}
%\comment{
We now give  a small-step semantics for our  language.
We first specify the domains used by the semantics:
\[
\renewcommand{\arraystretch}{1}
\begin{array}{@{}r@{\ }l@{\ \ }c@{\ \ }l@{\hspace{0.5em}}l}
\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow (Data + \{empty\}) & \mbox{-- Heap}\\
d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values \& Closures} \\
v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Data \times Data}& \mbox{-- Values}\\
c:   & \mathit{Clo} &=& \mathit{(App \times Env)}& \mbox{-- Closures}\\
\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} &
\mbox{-- Environment} \\
\end{array}
\]

@@

The  semantics  of   expressions  (and  applications\footnote{In  most
 contexts, we  shall use the  term 'expression' and the  notation $e$
  for both expressions and applications.}) 
are given by transitions of the form $\rho, \stk, \heap, e \rightsquigarrow
\rho',  \stk', \heap',  e'$.  Here  \stk\ is  a stack  of continuation
frames.  Each  continuation frame  is of the  form $(\rho,  \ell, e)$,
signifying that the  location $\ell$ has to be updated  with the value
of the  currently evaluating  expression, and $e$  is to  be evaluated
next  in the  environment $\rho$.  
%% Given a state $\rho, \stk, \heap,
%% e$  of  the  transition system, we shall call the pair $(\rho, e)$ as
%% the    \emph{current
%%   evaluation  context}, and for a continuation frame $(\rho',  \ell, e')$ in
%% \stk, we shall call the pair $(\rho', e')$ as a   
%% \emph{suspended  evaluation  context}.}
%% %% \warning{The
%% 3-tuple $\rho, \stk,  \heap, e$ representing the state  of the transition
%% system will be denoted $\mathcal{S}$.}
The  initial  state  of  the  transition  system  described  above  is
$([\;]_\rho,\,    (\rho_\mathit{init},   \, \ell_{\ans}  ,\,
(\print~\ans)):[\;]_{S}, 
[\;]_{\heap},\, (\mainpgm))$,  in which  $[\;]_\rho$,
$[\;]_{\mbox \footnotesize {H}}$
and $[\;]_{S}$ are    are the
empty environment, heap and stack respectively. The  initial stack consists
of  a single  continuation frame  in  which \ans\  is a  distinguished
variable  that  would   eventually  be  updated  with   the  value  of
(\mainpgm) and $\rho_\mathit{init}$ maps \ans\ to a location $\ell_{\ans}$.   In  addition, \print\ is  a function modeling  a printing
mechanism---a   standard   run-time   support  assumption   for   lazy
languages~\cite{Jones87}---that prints  the value of  (\mainpgm).  The operator $:$
pushes elements on top of the stack.

@@

%}\comment
%==============================================================
\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
%==============================================================
\input{fig-liveness}

@@

A variable is {\em live} if there  is a possibility of its value being
used in  future computations and  dead if  it is definitely  not used.
Heap-allocated data  needs a richer model  than classical liveness---a
model  which  talks  about  liveness of  references.   Using  $\acar$,
$\acdr$  to  represent access  using  $\CAR$  and $\CDR$  fields,  the
liveness of the structure reachable from a variable can be represented
by a set of {\em access paths} i.e.  prefix-closed set of strings from
$\{\acar,\acdr\}^\ast$.  Given an initial location $\ell$ (usually a 
reference corresponding to a variable) and  a heap  \heap, semantically  an access  path
$\alpha$ represents  a reference, denoted $\heap\llbracket\ell,  \alpha\rrbracket$, in
the heap  that is  obtained by  starting with  $\ell$ and  chasing the
\CAR\ or  \CDR\ fields in the heap as specified  by the  access path.  
$\heap\llbracket\ell, \alpha\rrbracket$  is defined only  if the path followed  in the
heap  is \emph{closure-free}  (does not  cross closures),  else it  is
undefined.

@@

@@

%\comment{  
Consider the  program in Figure~\ref{fig:mot-example2}.   As mentioned
earlier, a lazy evaluation of the $\LET$ expression at $\pi_1$ creates
a closure  for $(\length~\px)$  instead of  evaluating it.   Since the
closure may escape the scope in which it is created, it carries a copy
of $\px$ within itself.  We treat the  copy of $\px$ in the closure as
being separate from the $\px$ introduced  by the $\LET$, and call it a
\emph{closure  variable}.  For  liveness calculations,  such variables
are distinguished  from variables  introduced by  \LET s  and function
arguments  (called \emph{stack  variables}  since they  reside in  the
activation  stack). Notationally,  we distinguish  a closure  variable
from  its corresponding  stack variable  by subscripting  it with  the
program   point  where   the   closure  was   created\footnote{Mutiple
  occurrences  of the  same  variable in  an  application are  further
  distinguished by their positions in the application.}.

@@

The  variable \pz\   returned  at  $\psi_2$ depends  on  \py\ 
  through  the  expression $(+~\py~1)$.   $\py$  in  turn depends  on  the
  closure  variable  $\px_{\mathit{\pi}_1}$   through  $(\length~\px_{\mathit{\pi}_1})$.   We
  denote  this chain  of dependences  as $\lbrack  \psi_2\!\!\!:\!\!\!\pz \leftarrow
  (+~\py~1),\,  \py  \leftarrow   (\length~\px_{\mathit{\pi}_1})\rbrack$.   Indeed,  the
  closures  in  the  heap  are   a  runtime  representation  of  these
  dependences.  Since  $\pz$  is  evaluated   at  $\psi_2$  due  to  the
  expression $\SRETURN~\pz$,  the demand made by  the calling context(s)
  of  $f$ places  a demand  on  $\pz$ which  will impart  a liveness  to
  $\px_{\pi_1}$.  Other dependence chains which result in a liveness for
  $\px_{\pi_1}$       are       $\lbrack       \psi_1\!\!\!:\!\!\py       \leftarrow
  (\length~\px_{\pi_1})\rbrack$ and $\lbrack \psi_3\!\!\!:\!\!\!\pw \leftarrow (/~\pu~\py),\,\,
  \py  \leftarrow  (\length~\px_{\pi_1})\rbrack$.   The  liveness  analysis
  described in this section declares the liveness of $\px_{\pi_1}$ to be
  a union of the liveness arising  out of these dependence chains.  To
  be safe, a  garbage collection during evaluation of  $\py$ at $\psi_1$
  has  to  use  this  liveness  to copy  the  heap  from  $\px_{\pi_1}$.
  However,  notice that  if  a garbage  collection  takes place  while
  evaluating $\pz$ at $\psi_2$, it can safely consider only the liveness
  arising  out of  the dependence  chain\linebreak $\lbrack  \psi_2\!\!\!:\!\!\!\pz \leftarrow
  (+~\py~1),\,   \py  \leftarrow   (\length~\x_{\pi_1})\rbrack$.   The   garbage
  collection  scheme described  in Section~\ref{sec:GC-scheme}  uses a
  generalization  of   this  observation   to  dynamically   select  a
  evaluation point specific liveness in order to collect more garbage.

@@

Note that our proofs will be for a single round of minefield execution
i.e.  the evaluation of $(\mainpgm)$ (or $e_{\mainpgm}$, for call-less
programs) to  its WHNF driven  by the printing mechanism.   With minor
variations,  the   proof  will  also  be   applicable  for  subsequent
evaluations initiated by $\print$.
%% We show  next that the  minefield execution of a  call-less program
%% cannot go \bang.

@@

\begin{proof}
Assume to  the contrary that  a program $P$  enters the
\bang\  state.  \warning{We can  transform  $P$  to a  call-less
program $P'$ such that  the minefield executions of $P$
and $P'$  are identical  except for change  of variable
names. }             However,                 by
Lemma~\ref{lemma:call-less-cannot-go-bang} we know that
$P'$,   a   call-less   program,   cannot   enter   the
\bang\  state.  Therefor  $P$  also  cannot  enter  the
\bang\ state.
\end{proof}

@@

In general, the equations for \Lfonly\ are recursive as
in the  case of  \Lfonly$_\mathit{f}$. As  mentioned in
\cite{asati14lgc},   a   closed   form   solution   for
$\Lfone{\mathit{f}}$ can  be derived by  observing that
each of  the liveness rules  modifies a demand  only by
prefixing  it with  symbols  in  the alphabet  $\lbrace
\acar, \acdr,\bcar,  \bcdr, \clazy  \rbrace$. Therefore
we can  assume that $\Lf{f}{i}{\sigma}$ has  the closed
form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} = \Df{f}{i}\sigma
\end{eqnarray}
where \Df{f}{i}  are sets of strings  over the alphabet
mentioned above.  Substituting the  guessed form in the
equation   describing    \Lfonly$_{\mathit   f}$,   and
factoring  out   $\sigma$,  we  get  an   equation  for
\Df{f}{i}  that   is  independent  of   $\sigma$.   Any
solution   for   \Df{f}{i}   yields  a   solution   for
\Lfonly$_{\mathit  f}$.   Applied to  \Lfonly$_{\mathit
  \length}$, we get:
  \begin{eqnarray*}
&&  \Lf{\length }{1}{\sigma} = \Df{\length}{1}\sigma,~\mbox{and}\\
&&   \Df{\length}{1} = \clazy \cup \acdr\Df{\length}{1}\clazy
       \cup \clazy\Df{\length}{1}{\clazy}
  \end{eqnarray*}

@@

\begin{definition}\label{def:specialgrammar}
Consider the  grammar $(N,T,  p_1\cup p_2,S)$  in which
$N$ is  a set  of non-terminals,  $T =  \{\acar, \acdr,
\bcar,  \bcdr,  \clazy,  \$\}$,   $p_1$  is  a  set  of
context-free    productions     that    contains    the
distinguished  production   $S  \rightarrow  \alpha\$$,
$\alpha$ is a  string of grammar symbols  that does not
contain $S$, and $p_2$ is the fixed set of unrestricted
productions    $\bcar\acar    \rightarrow    \epsilon$,
$\bcdr\acdr    \rightarrow   \epsilon$,    $\clazy\acar
\rightarrow \clazy$,  $\clazy\acdr \rightarrow \clazy$,
and $\clazy\$ \rightarrow \epsilon$.
\end{definition}

@@

\begin{proof}
Given a Turing machine and  an input $w\in (1+0)^*$, we
construct a grammar $G$ such that the machine will halt
on    $w$    if    and   only    if    $\epsilon    \in
\mathscr{L}(G)$. The grammar includes  the fixed set of
unrestricted               productions               in
Definition~\ref{def:specialgrammar}.

@@

\begin{quote}
  If    the   configuration    of   the    TM   is
  $w_l(\state,c)w_r$, then the  sentential form will be
  $\overline{w}_l\nont{}{c}\,w_r$,                where
  $\overline{w}_l$ is  the same as $w_l$  but with each
  symbol $d$ in $w_l$ replaced by $\overline{d}$.
\end{quote}

@@

It is clear that if the TM given as an instance of the halting problem
accepts the input string, then the earlier derivation can be replayed,
every   time   replacing   $\nont{\mathit{\mathit{all}}}{}$   in   the
sentential form by $\epsilon$. However, if  the TM does not accept the
input string, then every sentential form derived from the start symbol
would    have     at    least    one    non-terminal from the grammar
of Lemma~\ref{lemma:grammar-undecidable} that is different
from $\nont{\mathit{final}}{c}$.   Thus  the  empty  string  would  not  be
derivable from the grammar.
\end{proof}

>>>>

%% We use the  result of the analysis to annotate  each potential garbage
%% collection  point  in   the  program  with  a   set  of  deterministic
%% finite-state automata (DFA) describing the liveness at that point.
To show the effectiveness of our method, we
implemented a copying collector that uses the results of the liveness
analysis to preserve live 
objects, both evaluated (i.e., in WHNF) and closures.  Our experiments
confirm  that  for  programs  running with  a  liveness-based  garbage
collector,   there  is   a   significant  decrease   in  peak   memory
requirements.   In  addition,   a  sizable  reduction  in   the  number  of
collections  ensures that  in spite  of using  a more  complex garbage
collector,  the execution  time  of  a program  remains comparable  with
the same program running with a reachability-based collector.
\warning{On our benchmark programs, we obtain a decrease of up to 75\%
  in the number of collections, and up to 50\% decrease in peak memory
  requirements.}
\end{abstract}

@@

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The  analysis  has  two parts.   We  first  calculate  a
context-sensitive  summary   of  each   function  as  a   {\em  demand
  transformer} that transforms a {\em  symbolic} demand on its body to
demands  on its  arguments.   This  summary is  used  to step  through
function calls  during analysis.   The concrete  demand on  a function
body  is  obtained through  a  conservative  approximation similar  to
0-CFA~\cite{Shivers:1988} that  combines the demands on  all the calls
to  the function.   The result  of the  analysis is  an annotation  of
certain program points with  deterministic finite-state automata (DFA)
capturing the  liveness of  variables at  these points.   Depending on
where garbage collection is triggered, the collector consults a set of
automata to restrict reachability during  marking.  This results in an
increase in  the garbage reclaimed  and consequently in  fewer garbage
collections.

While this idea has been shown  to be effective for a first-order {\em
  eager}  language~\cite{asati14lgc}, a  straightforward extension  of
the technique is not possible for lazy languages, where heap-allocated
objects may  include closures (runtime representations  of unevaluated
expressions).   Firstly, since  data  is made  live  by evaluation  of
closures, and  in lazy languages the  place in the program  where this
evaluation  takes  place  cannot be  statically  determined,  laziness
complicates  liveness analysis  itself.  Moreover,  for liveness-based
garbage collection to  be effective, we need to extend  it to closures
apart from  evaluated data.  Since a  closure can escape the  scope in
which it was created, it has  to carry the liveness information of its
free variables. As execution progresses and possible future
uses are eliminated,  we update the liveness information  in a closure
with a  more precise version.  Because of these reasons,   the garbage
collector  also   becomes  significantly   more  complicated   than  a
liveness-based collector for a eager language.

Experiments  with  a   single  generation  copying  collector
  (Section~\ref{sec:experiments})  confirm  the  expected  performance
  benefits.   Liveness-based  collection  results in  an  increase  in
  garbage reclaimed.   As a  consequence, there is  a reduction  in the
  number of collections from \warning{1.6X to 23X} and a decrease in the
  minimum memory requirement from  \warning{1X to 1389X}. 
  As an added benefit, there is also a reduction in the overall
  execution time in 5 out of 12 benchmark programs.

@@

$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, \heap,\, (+~x~y)
\rightsquigarrow
\rho',\, (\rho, \rho(x), (+~x~y))\!:\!S,\, \heap,\, s$      &
{\sc prim-1-clo} \\
\hline
$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho,\, S,\, \heap,\, (+~x~y)
\rightsquigarrow
\rho', (\rho,\, \rho(y),\, (+~x~y))\!:\!S,\, \heap,\, s$      &
{\sc prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho,\, S,\, \heap,\,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})],\, S,\, \heap,\, e_{\mathit{f}}$      &
{\sc funcall} \\
\hline
$\ell$ is a new location& {$\rho,\, S,\, \heap,\, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell],\, S,\, \heap[\ell := \langle s,
    \lfloor\rho\rfloor_{FV(s)} %% \oplus [x \mapsto  \ell] -- NO LAZY LETS
    \rangle],\, e$} &
{\sc let} \\
\hline
$\heap(\rho(x)) \ne 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho,\, S,\, \heap,\,  e_1$ & {\sc if-true} \\
\hline
$\heap(\rho(x)) = 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho,\, S, \,\heap, \, e_2$ & {\sc if-false} \\
\hline
$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho',\, (\rho, \rho(x), (\SIF~x~e_1~e_2))\!:\!S,\, \heap, \, s$}
&
{\sc if-clo} \\
\hline
{$\heap(\rho(x))~\mbox{is in WHNF with value}~v$}& $\rho,\, (\rho', \ell,
e)\!:\!S,\, \heap,\,
(\SRETURN~x)  \rightsquigarrow \rho',\, S,\, \heap[\ell := v],\, e$ &
{\sc return-whnf}\\
\hline
$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\, (\SRETURN~x)
  \rightsquigarrow$
$\rho',\, (\rho, \rho(x), (\SRETURN~x))\!:\!S,\, \heap,\,  s$} &
{\sc return-clo} \\
\hline
\end{tabular}
\caption{The small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
%}\comment
\vspace*{-6pt}
\end{figure*}

\subsection{Semantics}\label{sec:semantics}
%\comment{
We now give  a small-step semantics for our  language.
We first specify the domains used by the semantics:
\[
\renewcommand{\arraystretch}{1}
\begin{array}{@{}r@{\ }l@{\ \ }c@{\ \ }l@{\hspace{0.5em}}l}
\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow (Data + \{empty\}) & \mbox{-- Heap}\\
d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values \& Closures} \\
v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Data \times Data}& \mbox{-- Values}\\
c:   & \mathit{Clo} &=& \mathit{(App \times Env)}& \mbox{-- Closures}\\
\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} &
\mbox{-- Environment} \\
\end{array}
\]

@@

The  semantics  of   expressions  (and  applications\footnote{In  most
 contexts, we  shall use the  term 'expression' and the  notation $e$
  for both expressions and applications.}) 
are given by transitions of the form $\rho, \stk, \heap, e \rightsquigarrow
\rho',  \stk', \heap',  e'$.  Here  \stk\ is  a stack  of continuation
frames.  Each  continuation frame  is of the  form $(\rho,  \ell, e)$,
signifying that the  location $\ell$ has to be updated  with the value
of the  currently evaluating  expression, and $e$  is to  be evaluated
next  in the  environment $\rho$.  
%% Given a state $\rho, \stk, \heap,
%% e$  of  the  transition system, we shall call the pair $(\rho, e)$ as
%% the    \emph{current
%%   evaluation  context}, and for a continuation frame $(\rho',  \ell, e')$ in
%% \stk, we shall call the pair $(\rho', e')$ as a   
%% \emph{suspended  evaluation  context}.}
%% %% \warning{The
%% 3-tuple $\rho, \stk,  \heap, e$ representing the state  of the transition
%% system will be denoted $\mathcal{S}$.}
The  initial  state  of  the  transition  system  described  above  is
\[([\;]_\rho,\,     (\rho_\mathit{init},     \,    \ell_{\ans}     ,\,
(\print~\ans)):[\;]_{S},   [\;]_{\heap},\,  (\mainpgm))\]   in  which
$[\;]_\rho$, $[\;]_{\mbox  \footnotesize {H}}$ and $[\;]_{S}$  are are
the empty environment, heap and  stack respectively. The initial stack
consists  of  a  single  continuation   frame  in  which  \ans\  is  a
distinguished variable that would eventually be updated with the value
of  (\mainpgm)  and  $\rho_\mathit{init}$  maps \ans\  to  a  location
$\ell_{\ans}$.  In addition, \print\ is a function modeling a printing
mechanism---a   standard   run-time   support  assumption   for   lazy
languages~\cite{Jones87}---that prints  the value of  (\mainpgm).  The
operator~$:$ pushes elements on top of the stack.

@@

@@

A variable is {\em live} if there  is a possibility of its value being
used in  future computations and  dead if  it is definitely  not used.
Heap-allocated data  needs a richer model  than classical liveness---a
model which talks about liveness of references.  An {\em access path}
is a  prefix-closed set of strings  over $\{\acar,\acdr\}^\ast$, where
$\acar$,  $\acdr$ represent  access  using $\CAR$  and $\CDR$  fields.
Given an initial location $\ell$ (usually a reference corresponding to
a variable)  and a  heap \heap, semantically  an access  path $\alpha$
represents     a      reference,     denoted     $\heap\llbracket\ell,
\alpha\rrbracket$,  in the  heap  that is  obtained  by starting  with
$\ell$ and chasing the \CAR\ or  \CDR\ fields in the heap as specified
by  the  access   path.   $\heap\llbracket\ell,  \alpha\rrbracket$  is
defined only if  the path followed in the  heap is \emph{closure-free}
(does not cross closures), else it is undefined.

@@

%==============================================================
\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
%==============================================================
\input{fig-liveness}

@@

Consider the  program in Figure~\ref{fig:mot-example2}.   As mentioned
earlier, a lazy evaluation of the $\LET$ expression at $\pi_1$ creates
a closure  for $(\length~\px)$  instead of  evaluating it.   Since the
closure may escape the scope in which it is created, it carries a copy
of $\px$ within itself.  We treat the  copy of $\px$ in the closure as
being separate from the $\px$ introduced  by the $\LET$, and call it a
\emph{closure  variable}.  For  liveness calculations,  such variables
are distinguished  from variables  introduced by  \LET s  and function
arguments  (called \emph{stack  variables}  since they  reside in  the
activation  stack). Notationally,  we distinguish  a closure  variable
from  its corresponding  stack variable  by subscripting  it with  the
program   point  where   the   closure  was   created\footnote{Multiple
  occurrences  of the  same  variable in  an  application are  further
  distinguished by their positions in the application.}.

@@

The  variable \pz\   returned  at  $\psi_3$ depends  on  \py\ 
  through  the  expression $(+~\py~1)$.   $\py$  in  turn depends  on  the
  closure  variable  $\px_{\mathit{\pi}_1}$   through  $(\length~\px_{\mathit{\pi}_1})$.   We
  denote  this chain  of dependences  as $\lbrack  \psi_3\!\!\!:\!\!\!\pz \leftarrow
  (+~\py~1),\,  \py  \leftarrow   (\length~\px_{\mathit{\pi}_1})\rbrack$.   Indeed,  the
  closures  in  the  heap  are   a  runtime  representation  of  these
  dependences.  Since  $\pz$  is  evaluated   at  $\psi_3$  due  to  the
  expression $\SRETURN~\pz$,  the demand made by  the calling context(s)
  of  $f$ places  a demand  on  $\pz$ which  will impart  a liveness  to
  $\px_{\pi_1}$.  Other dependence chains which result in a liveness for
  $\px_{\pi_1}$       are       $\lbrack       \psi_1\!\!\!:\!\!\py       \leftarrow
  (\length~\px_{\pi_1})\rbrack$ and $\lbrack \psi_2\!\!\!:\!\!\!\pw \leftarrow (/~\pu~\py),\,\,
  \py  \leftarrow  (\length~\px_{\pi_1})\rbrack$.   The  liveness  analysis
  described in this section declares the liveness of $\px_{\pi_1}$ to be
  a union of the liveness arising  out of these dependence chains.  To
  be safe, a  garbage collection during evaluation of  $\py$ at $\psi_1$
  has  to  use  this  liveness  to copy  the  heap  from  $\px_{\pi_1}$.
  However,  notice that  if  a garbage  collection  takes place  while
  evaluating $\pz$ at $\psi_3$, it can safely consider only the liveness
  arising  out of  the dependence  chain\linebreak $\lbrack  \psi_3\!\!\!:\!\!\!\pz \leftarrow
  (+~\py~1),\,   \py  \leftarrow   (\length~\x_{\pi_1})\rbrack$.   The   garbage
  collection  scheme described  in Section~\ref{sec:GC-scheme}  uses a
  generalization  of   this  observation   to  dynamically   select  a
  evaluation point specific liveness in order to collect more garbage.

@@

Note that our proofs will be for a single round of minefield execution
i.e.  the evaluation of $(\mainpgm)$ (or $e_{\mainpgm}$, for call-less
programs)  to   its  WHNF  driven  by   the  printing  mechanism~(Section
\ref{sec:semantics}).  With  minor variations,  the proof will  also be
applicable for subsequent evaluations initiated by $\print$.
%% We show  next that the  minefield execution of a  call-less program
%% cannot go \bang.

@@

\begin{proof}
Assume to  the contrary that  a program $P$  enters the
\bang\  state.  \warning{We can  transform  $P$  to a  call-less
program $P'$ such that  the minefield executions of $P$
and $P'$  are identical  except for change  of variable
names. }             However,                 by
Lemma~\ref{lemma:call-less-cannot-go-bang} we know that
$P'$,   a   call-less   program,   cannot   enter   the
\bang\  state.  Therefore   $P$  also  cannot  enter  the
\bang\ state.
\end{proof}

@@

In general, the equations for \Lfonly\ are recursive as in the case of
\Lfonly$_\mathit{f}$. A closed  form solution for $\Lfone{\mathit{f}}$
can be derived by observing that each of the liveness rules modifies a
demand  only by  prefixing it  with symbols  in the  alphabet $\lbrace
\acar, \acdr,\bcar,  \bcdr, \clazy  \rbrace$. Therefore we  can assume
that $\Lf{f}{i}{\sigma}$ has the closed form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} = \Df{f}{i}\sigma
\end{eqnarray}
where \Df{f}{i}  are sets of strings  over the alphabet
mentioned above.  Substituting the  guessed form in the
equation   describing    \Lfonly$_{\mathit   f}$,   and
factoring  out   $\sigma$,  we  get  an   equation  for
\Df{f}{i}  that   is  independent  of   $\sigma$.   Any
solution   for   \Df{f}{i}   yields  a   solution   for
\Lfonly$_{\mathit  f}$.   Applied to  \Lfonly$_{\mathit
  \length}$, we get:
  \begin{eqnarray*}
&&  \Lf{\length }{1}{\sigma} = \Df{\length}{1}\sigma,~\mbox{and}\\
&&   \Df{\length}{1} = \clazy \cup \acdr\Df{\length}{1}\clazy
       \cup \clazy\Df{\length}{1}{\clazy}
  \end{eqnarray*}

@@

\begin{definition}\label{def:specialgrammar}
Consider the  grammar $(N,T,  p_1\cup p_2,S)$  in which
$N$ is  a set  of non-terminals,  $T =  \{\acar, \acdr,
\bcar,  \bcdr,  \clazy,  \$\}$,   $p_1$  is  a  set  of
context-free    productions     that    contains    the
distinguished  production   $S  \rightarrow  \alpha\$$,
$\alpha$ is a  string of grammar symbols  that does not
contain $S$, and $p_2$ is the fixed set of unrestricted
productions    $\bcar\acar    \rightarrow    \epsilon$,
$\bcdr\acdr    \rightarrow   \epsilon$,    $\clazy\acar
\rightarrow \clazy$,  $\clazy\acdr \rightarrow \clazy$,
and $\clazy\$ \rightarrow \$ $.
\end{definition}

@@

\begin{proof}
Given a Turing machine and  an input $w\in (1+0)^*$, we
construct a grammar $G$ such that the machine will halt
on    $w$    if    and   only    if    $\$    \in
\mathscr{L}(G)$. The grammar includes  the fixed set of
unrestricted               productions               in
Definition~\ref{def:specialgrammar}.

@@

\begin{quote}
  If    the   configuration    of   the    TM   is
  $w_l(\state,c)w_r$, then the  sentential form will be
  $\overline{w}_l\nont{}{c}\,w_r\$ $,                where
  $\overline{w}_l$ is  the same as $w_l$  but with each
  symbol $d$ in $w_l$ replaced by $\overline{d}$.
\end{quote}

@@

It is clear that if the TM given as an instance of the halting problem
accepts the input string, then the earlier derivation can be replayed,
every   time   replacing   $\nont{\mathit{\mathit{all}}}{}$   in   the
sentential form by $\epsilon$. However, if  the TM does not accept the
input string, then every sentential form derived from the start symbol
would    have     at    least    one    non-terminal from the grammar
of Lemma~\ref{lemma:grammar-undecidable} that is different
from $\nont{\mathit{final}}{c}$.   Thus  $\$$  would  not  be
derivable from the grammar.
\end{proof}
